---
title: "Additional notes on EDA"
subtitle: "Math 275, Winter 2018"
author: "Adam Loy"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: center, middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# Distributional Assessment

---

# Q-Q plots

**Goal:** Compare two sets of quantiles to see if the distributions could be the same
  
  + sample vs. sample
  
  + **sample vs. theoretical**

---

class: center, middle

# Build your intuition
### http://gosset.wharton.upenn.edu:3838/foster/quantile_slider/

(Select "Sample distribution?")

---

# Example 1

Could the sample plausibly follow a normal distribution?

<img src="http://aloy.rbind.io/math275/slides/figs/example-heavy-tail.pdf" alt="some text"  width="400" height="400">


---


# Example 2

Could the sample plausibly follow a normal distribution?

<img src="http://aloy.rbind.io/math275/slides/figs/qqplot1.pdf" alt="some text"  width="400" height="400">


---

# Example 3

Could the sample plausibly follow a normal distribution?

<img src="http://aloy.rbind.io/math275/slides/figs/example-right-skew.pdf" alt="some text"  width="400" height="400">

---

# Take aways

- If a sample plausibly follows a distribution, then the Q-Q plot will be roughly diagonal

- Don't worry about "minor wiggles"

- Q-Q plots can be hard to interpret for small sample sizes

```{r echo=FALSE, fig.height=5, fig.width=14}
seed <- c(1, 2, 99999)
par(mfrow = c(1,3), cex = 1.1)
for(i in seed) {
  set.seed(i)
  x <- rnorm(10)
  qqnorm(x, ylim = c(-2,2), xlim = c(-2,2), pch = 16); qqline(x)
}
```

---

# Best practices

1. Experiment with histogram binwidth/breaks

1. Bar charts start at zero

1. Use histograms, not pie charts

1. Label your axes, including your units

1. Give context in a title or caption

1. Be able to describe every graph that you use

